---
title: "Problem sheet: clustering using k-means"
output: html_notebook
---

In this problem set, we are going to write our own k-means clustering algorithm. The dataset we are going to use is the famous (Fisher's or Anderson's) iris data set, which gives the measurements in centimeters of the variables sepal length and width and petal length and width, respectively, for 50 flowers from each of 3 species of iris. The species are Iris setosa, versicolor, and virginica.

1. Load the iris data and examine the data. Here we are not going to use numerical data on sepal length, sepal width, petal length and petal width to cluster the specimens.
```{r}
rm(list=ls())
library(tidyverse)
library(reshape2)

iris
x <- iris %>% 
  select(-Species)
```


2. Write a function which samples $k$ data points (without replacement) to be the initial centroids.
```{r}
f_choose_random_centroids <- function(x, k=3) {
  idxs <- 1:nrow(x)
  idxs <- sample(idxs, k)
  x[idxs, ]
}
```

3. Write a function which calculates the Euclidean distance between two points.
```{r}
f_eucl_dist <- function(x1, x2) {
  sqrt(sum((x1 - x2)^2))
}
```

4. Write a function which finds the distance between each row in a matrix (where each of its rows is a data point) and a centroid.
```{r}
f_eucl_dist_all <- function(x, centroid) {
  map_dbl(seq(1, nrow(x), 1), ~f_eucl_dist(centroid, x[., ]))
}
```

5. Write a function which finds the distance between each row in a matrix (where each of its rows is a data point) and a list of centroids. It may be easiest to return these distances in a matrix, where each row corresponds to a data point and each column to a centroid.
```{r}
f_eucl_dist_centroids <- function(x, centroids) {
  m_dist <- matrix(nrow = nrow(x), ncol = nrow(centroids))
  for(i in 1:nrow(centroids))
    m_dist[, i] <- f_eucl_dist_all(x, centroids[i, ])
  m_dist
}
```

6. Write a function which assigns each data point to a cluster according to its nearest centroid. The function should return a list of centroid ids: one per each data point.
```{r}
f_cluster <- function(x, centroids) {
  m_dist <- f_eucl_dist_centroids(x, centroids)
  cluster_id <- vector(length = nrow(m_dist))
  for(i in seq_along(cluster_id))
    cluster_id[i] <- which.min(m_dist[i, ])
  cluster_id
}
```

7. Create a function which calculates new centroids using the points which have been assigned to each cluster.
```{r}
f_recalculate_centroids <- function(cluster_ids, x) {
  df <- x %>% 
    as.data.frame() %>% 
    mutate(cluster=cluster_ids)
  mu <- df %>% 
    group_by(cluster) %>% 
    summarise_all(.funs=mean) %>% 
    select(-cluster)
  mu
}
```

8. Putting your functions together in sequence create a function that does k-means clustering! To do this, do the following:

create a for loop where:

- if first time round loop, choose random centroids to be data points from the dataset; if subsequent, calculate centroids using your function that determines the centroids of each cluster
- calculates distance of all points from all centroids, and assigns clusters to the points based on their minimal distance

For simplicitly, here we terminate the loop after 50 iterations.
```{r}
# here I just use 50 iterations as a stopping criteria
# better to look at cluster identities and only stop
# once these stop changing
f_kmeans <- function(x, k, niter=50) {
  for(i in 1:niter) {
    if(i == 1)
      centroids <- f_choose_random_centroids(x, k)
    else
      centroids <- f_recalculate_centroids(cluster_ids, x)
    cluster_ids <- f_cluster(x, centroids)
  }
  cluster_ids
}
```

9. Using $k=3$ clusters, use your k-means clustering function to determine cluster ids.
```{r}
clusters <- f_kmeans(x, 3)
```

10. How do your clusters correspond to species?
```{r}
df <- iris %>% 
  as.data.frame() %>% 
  mutate(cluster=clusters)
table(df$Species, df$cluster)
```
Does a pretty good job of separating out the three species. Setosa is more different from the other two classes.

11. What happens if you choose only two clusters?
```{r}
clusters <- f_kmeans(x, 2)
df <- iris %>% 
  as.data.frame() %>% 
  mutate(cluster=clusters)
table(df$Species, df$cluster)
```
Splits data in setosa and others. This makes sense, since assuming 3 clusters mixtured up the versicolor and virginica datasets
